---
title: "3_models"
author: "Jennifer Neumaier"
date: "3/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import data
```{r}
source("./src/convenience.R")
source("./src/data_analysis.R")
```


Libraries
```{r}
library("tidyverse")
library("tidymodels")
library("mikropml")

# https://cran.r-project.org/web/packages/easyCODA/easyCODA.pdf
library("easyCODA")

# https://htmlpreview.github.io/?https://github.com/malucalle/selbal/blob/master/vignettes/vignette.html
# devtools::install_github(repo = "malucalle/selbal")
library("selbal")

# https://cran.r-project.org/web/packages/codacore/codacore.pdf
library("codacore")
```

## Classfication

Import data and merge with metadata
```{r}
# Import data
file_path <- "./data/PCOS"
data_transformed <- import(path = file_path, pattern = ".*(transform_*).*", header = TRUE)
data_imputed <- import(path = file_path, pattern=".*(abundances_.*).*", header = TRUE)
metadata <- read.table("./data/PCOS/PCOS_metadata.txt", header = TRUE)

# merge data_transformed
for (i in 1:length(data_transformed)){
  data_transformed[[i]] <- data_transformed[[i]] %>% 
  left_join(metadata, by = SampleID) %>% 
  relocate(c(PCOS_Riikka, BMI), .after = SampleID) %>% 
  dplyr::select(-c(SampleID, PCOS_Riikka))
}


# merge data_imputed and also split
train_data <- list()
test_data <- list()

for (i in 1:length(data_imputed)){
  
  # merge PCOS_data with metadata
  data_imputed[[i]] <- data_imputed[[i]] %>% 
  left_join(metadata, by = "SampleID") %>% 
  relocate(c(PCOS_Riikka, BMI), .after = SampleID) %>% 
  dplyr::select(-c(SampleID, PCOS_Riikka))
  
  # Split the data
  set.seed(2022)
  data_split <- initial_split(data_imputed[[i]], prop = 0.8, strata = BMI) 
  train_data[[i]] <- training(data_split) %>% as.data.frame()
  test_data[[i]] <- testing(data_split) %>% as.data.frame()
  
}

names(train_data) <- names(data_imputed)
names(test_data) <- names(data_imputed)

# save integers that control training set
idx <- data_split$in_id
```

mikropml
```{r}
model_name <- "glmnet"

training_AUC <- list()
test_AUC <- list()
counter <- seq(1,2,1)

# go through each data set in data_list
for (i in 1:length(data_transformed)) { 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  # repeat experiment n times
  for (j in counter) { 
    
    ml_result <- run_ml(data_transformed[[i]],
                     model_name,
                     outcome_colname = predictor, 
                     kfold = 5,
                     cv_times = 10,
                     training_frac = train_ids, # training fraction taken from split for codacore
                     seed = NA)
    
    # save training AUC in lists
    training_res[j] <- ml_result$performance[1]
    test_res[j] <- ml_result$performance[3]
  }
  
  # save lists in final list
  training_AUC[[i]] <- training_res
  test_AUC[[i]] <- test_res
}
```

Codacore
```{r}
coda_training_AUC <- list()
coda_test_AUC <- list()

counter <- seq(1,5,1)

for (i in 1:length(train_data)) { 
  
  # generate features (x) and variable (y) for train
  x_train <- train_data[[i]]
  x_train <- x_train[, 2:ncol(x_train)]
  
  y_train <- train_data[[i]]
  y_train <- y_train[,1] 
  
  # generate features (x) and variable (y) for test
  x_test <- test_data[[i]]
  x_test <- x_test[, 2:ncol(x_test)]
  
  y_test <- test_data[[i]]
  y_test <- y_test[,1] 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  for (j in counter) {
    # models
    model_bal_0 <- codacore(
      x_train, 
      y_train, 
      objective = "regression",
      logRatioType = "balances",
      cvParams = c(numFolds = 5),
      lambda = 0,
      verbose = FALSE, 
      fast = TRUE
    )
    
    model_bal_1 <- codacore(
      x_train, 
      y_train,
      objective = "regression",
      logRatioType = "balances",  
      cvParams = c(numFolds = 5),
      lambda = 1,
      verbose = FALSE, 
      fast = TRUE    
    )
    
      tryCatch(
        expr = { 
          
          # save training RMSE in lists
          training_res[[j]] <- c(as.numeric(model_bal_0$ensemble[[1]]$RMSE), 
                                 as.numeric(model_bal_1$ensemble[[1]]$RMSE))
          
          # get test RMSE by predicting manually
          pred_bal_0_lr1 <- predict(model_bal_0, x_test, logits = FALSE, numLogRatios = 1)
          pred_bal_0 <- predict(model_bal_0, x_test, logits = FALSE)
          
          pred_bal_1_lr1 <- predict(model_bal_1, x_test, logits = FALSE, numLogRatios = 1)
          pred_bal_1 <- predict(model_bal_1, x_test, logits = FALSE)
          
          
          # test_res[[j]] <- c(as.numeric(sqrt(mean((y_test - pred_bal_0_lr1)^2))), 
          #                    as.numeric(sqrt(mean((y_test - pred_bal_0)^2))), 
          #                    as.numeric(sqrt(mean((y_test - pred_bal_1_lr1)^2))),
          #                    as.numeric(sqrt(mean((y_test - pred_bal_1)^2))))
          
          
          test_res[[j]] <- c(as.numeric(ModelMetrics::rmse(actual = y_test, predicted = pred_bal_0_lr1, quiet = T)), 
                             as.numeric(ModelMetrics::rmse(actual = y_test, predicted = pred_bal_0, quiet = T)), 
                             as.numeric(ModelMetrics::rmse(actual = y_test, predicted = pred_bal_1_lr1, quiet = T)),
                             as.numeric(ModelMetrics::rmse(actual = y_test, predicted = pred_bal_1, quiet = T)))
        }
      )
  }
  
  # save lists in final list
  coda_training_AUC[[i]] <- training_res
  coda_test_AUC[[i]] <- test_res

} 

coda_training <- coda_training_AUC
coda_test <- coda_test_AUC

```

plot mikropml
```{r}
set_names <- names(data_transformed)

names(training_AUC) <- set_names
names(test_AUC) <- set_names

training_df <- stack(training_AUC)
test_df <- stack(test_AUC)

training_df <- training_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "training")

test_df <- test_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "test")
  
final_df_AUC <- rbind(training_df, test_df)

graph <- ggplot(final_df_AUC, aes(x = Dataset, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  labs(title = "AUC over various transformations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graph

ggsave(plot = graph, filename = "./figs/PCOS_glmnet_same_split.jpeg", device = "jpeg", height = 6, width = 7)
```

plot codacore
coda_test_AUC:  1. pred_bal_0_lr1, 2. pred_bal_0, 3. pred_bal_1_lr1; 4. pred_bal_1
coda_training_AUC: 1. model_bal_0 AUC, 2. model_bal_1 AUC

same format as mikropml:
- x-axis: 4 models for each PCOS_10 and PCOS_50 with e.g. model_bal_0 for both pred_bal_0_lr1 and pred_bal_0
```{r}
# set names
set_names <- names(data_imputed)

names(coda_training_AUC) <- set_names
names(coda_test_AUC) <- set_names

# extract first element of coda_training AUC
df_final <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "PCOS_abundances_10_pc")

df <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)


# same for PCOS_50
df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC
df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "PCOS_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC PCOS_50
df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "PCOS_abundances_50_pc")

df_final <- rbind(df_final, df)


# plot
graph <- ggplot(df_final, aes(x = model, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  facet_wrap(~dataset, scale="free") + 
  labs(title = "AUC CoDaCoRe") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

graph
ggsave(plot = graph, filename = "./figs/PCOS_codacore_same_split.jpeg", device = "jpeg", height = 6, width = 7)

```

## Regression

Import data and merge with metadata
```{r}
# Import data
data_transformed <- import(path = "./data/transformed/pseudocount", pattern="PCOS_\\d*", header = TRUE)
data_imputed <- import(path = "./data/processed", pattern="PCOS_abundances_\\d*_pc", header = TRUE)
metadata <- read.table("./data/processed/PCOS_metadata.txt", header = TRUE)

# merge data_transformed
for (i in 1:length(data_transformed)){
  data_transformed[[i]] <- data_transformed[[i]] %>% 
  left_join(metadata, by = "SampleID") %>% 
  relocate(c(PCOS_Riikka, BMI), .after = SampleID) %>% 
  select(-c(SampleID, BMI))
}

# merge data_imputed and also split
train_data <- list()
test_data <- list()

for (i in 1:length(data_imputed)){
  
  # merge PCOS_data with metadata
  data_imputed[[i]] <- data_imputed[[i]] %>% 
  left_join(metadata, by = "SampleID") %>% 
  relocate(c(PCOS_Riikka, BMI), .after = SampleID) %>% 
  select(-c(SampleID, BMI))
  
  # Split the data
  set.seed(2022)
  data_split <- initial_split(data_imputed[[i]], prop = 0.8, strata = PCOS_Riikka) 
  train_data[[i]] <- training(data_split) %>% as.data.frame()
  test_data[[i]] <- testing(data_split) %>% as.data.frame()
  
}

names(train_data) <- names(data_imputed)
names(test_data) <- names(data_imputed)

# save integers that control training set
idx <- data_split$in_id
```

mikropml
```{r}
training_AUC <- list()
test_AUC <- list()
counter <- seq(1,10,1)

# go through each data set in data_list
for (i in 1:length(data_transformed)) { 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  # repeat experiment n times
  for (j in counter) { 
    
    ml_result <- run_ml(data_transformed[[i]],
                     'glmnet',
                     outcome_colname = "PCOS_Riikka", 
                     kfold = 5,
                     cv_times = 10,
                     training_frac = idx, # training fraction taken from split for codacore
                     seed = NA)
    
    # save training AUC in lists
    training_res[j] <- ml_result$performance[1]
    test_res[j] <- ml_result$performance[3]
  }
  
  # save lists in final list
  training_AUC[[i]] <- training_res
  test_AUC[[i]] <- test_res
}
```

Codacore
```{r}
coda_training_AUC <- list()
coda_test_AUC <- list()

counter <- seq(1,10,1)

for (i in 1:length(train_data)) { 
  
  # generate features (x) and variable (y) for train
  x_train <- train_data[[i]]
  x_train <- x_train[, 2:ncol(x_train)]
  
  y_train <- train_data[[i]]
  y_train <- y_train[,1] 
  
  # generate features (x) and variable (y) for test
  x_test <- test_data[[i]]
  x_test <- x_test[, 2:ncol(x_test)]
  
  y_test <- test_data[[i]]
  y_test <- y_test[,1] 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  for (j in counter) {
    # models
    model_bal_0 <- codacore(
      x_train, 
      y_train, 
      objective = "binary classification",
      logRatioType = "balances",
      cvParams = c(numFolds = 5),
      lambda = 0,
      verbose = FALSE, 
      fast = TRUE
    )
    
    model_bal_1 <- codacore(
      x_train, 
      y_train,
      objective = "binary classification",
      logRatioType = "balances",  
      cvParams = c(numFolds = 5),
      lambda = 1,
      verbose = FALSE, 
      fast = TRUE    
    )
    
    # save training AUC in lists
    training_res[[j]] <- c(as.numeric(model_bal_0$ensemble[[1]]$AUC), 
                         as.numeric(model_bal_1$ensemble[[1]]$AUC))
    
    # get test AUC by predicting manually
    
    
    
    pred_bal_0_lr1 <- predict(model_bal_0, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_0 <- predict(model_bal_0, x_test, logits = FALSE)
    
    pred_bal_1_lr1 <- predict(model_bal_1, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_1 <- predict(model_bal_1, x_test, logits = FALSE)
    
    test_res[[j]] <- c(as.numeric(pROC::auc(y_test, pred_bal_0_lr1, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_0, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_1_lr1, quiet = T)),
                    as.numeric(pROC::auc(y_test, pred_bal_1, quiet = T)))
  }
  
  # save lists in final list
  coda_training_AUC[[i]] <- training_res
  coda_test_AUC[[i]] <- test_res

} 

```


plot mikropml
```{r}
set_names <- names(data_transformed)

names(training_AUC) <- set_names
names(test_AUC) <- set_names

training_df <- stack(training_AUC)
test_df <- stack(test_AUC)

training_df <- training_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "training")

test_df <- test_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "test")
  
final_df_AUC <- rbind(training_df, test_df)

graph <- ggplot(final_df_AUC, aes(x = Dataset, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  labs(title = "AUC over various transformations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graph

ggsave(plot = graph, filename = "./figs/PCOS_glmnet_same_split.jpeg", device = "jpeg", height = 6, width = 7)
```


plot codacore
coda_test_AUC:  1. pred_bal_0_lr1, 2. pred_bal_0, 3. pred_bal_1_lr1; 4. pred_bal_1
coda_training_AUC: 1. model_bal_0 AUC, 2. model_bal_1 AUC

same format as mikropml:
- x-axis: 4 models for each PCOS_10 and PCOS_50 with e.g. model_bal_0 for both pred_bal_0_lr1 and pred_bal_0
```{r}
# set names
set_names <- names(data_imputed)

names(coda_training_AUC) <- set_names
names(coda_test_AUC) <- set_names

# extract first element of coda_training AUC
df_final <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for PCOS_50
df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC
df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC PCOS_50
df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)


# plot
graph <- ggplot(df_final, aes(x = model, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  facet_wrap(~dataset, scale="free") + 
  labs(title = "AUC CoDaCoRe") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

graph
ggsave(plot = graph, filename = "./figs/CRC_codacore_same_split.jpeg", device = "jpeg", height = 6, width = 7)

```


## Try selbal with PCOS

selbal
https://github.com/malucalle/selbal
Additional parameters of selbal.cv function are:

    n.fold : a numeric value indicating the number of folds in the cross - validation procedure. When the response variable is dichotomous the cross-validation is performed so that the total proportion of cases and controls is preserved in each fold. Default n.fold = 5.
    n.iter: a numeric value indicating the number of iterations to implement; that is, the number of times that the cross - validation process is repeated. Default n.iter = 10.
    logit.acc: for dichotomous responses, the method to compute the association parameter. Default it is the AUC- ROC value (logit.acc = "auc"), but other alternatives are also included: logit.acc = "Dev" for the proportion of explained deviance ,logit.acc = "rsq" or logit.acc = "tjur" (see the paper to obtain more information);.
    maxV: the maximum number of variables to be included in a balance. Default maxV = 20.
    zero.rep : it defines the method used for the zero-replacement. If it is "bayes" the Bayesian - Multiplicative treatment implemented in {zCompositions} is applied. If it is "one", a pseudocount of 1 is added to the whole matrix.
    opt.cri parameter for selecting the method to determine the optimal number of variables. "max" to define this number as the number of variables which maximizes the association value or "1se" to take also the standard error into account. Default opt.cri = "1se".
    user_numVar: the number of variables to be included in a balance despite of the computed optimal value. Default user_numVar = NULL.

```{r}
PCOS_abundances_10 <- read.table("./data/processed/PCOS_abundances_10.txt", header = TRUE)
PCOS_metadata <- read.table("./data/processed/PCOS_metadata.txt", header = TRUE)

data_set <- PCOS_abundances_10 %>% 
  left_join(PCOS_metadata, by = "SampleID") %>% 
  relocate(c(PCOS_Riikka, BMI), .after = SampleID) %>% 
  select(-c(SampleID, BMI)) # change parameter here to decide which parameter should be the predictor (phenotype/BMI?)

x <- data_set[, 4:ncol(data_set)]
y <- ifelse(data_set[,2] == "control", 0, 1) # predictor column needs to be binary or continuous

selbal_results <- selbal.cv(x = x, y = y, n.fold = 5, n.iter = 10, logit.acc = "auc", zero.rep = "bayes")

?selbal.cv
selbal(x,y)
```



