---
title: "3_pipeline"
author: "Jennifer Neumaier"
date: "2/28/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pipeline

```{r}
source("./src/convenience.R")
source("./src/data_analysis.R")
```

Libraries
```{r}
library("tidyverse")
library("tidymodels")
library("mikropml")

# https://cran.r-project.org/web/packages/easyCODA/easyCODA.pdf
library("easyCODA")

# https://htmlpreview.github.io/?https://github.com/malucalle/selbal/blob/master/vignettes/vignette.html
# devtools::install_github(repo = "malucalle/selbal")
library("selbal")

# https://cran.r-project.org/web/packages/codacore/codacore.pdf
library("codacore")
```

## Classification

Import data and merge with metadata
```{r}
# Import data
data_transformed <- import(path = "./data/transformed/pseudocount", pattern="CRC_\\d*", header = TRUE)
data_imputed <- import(path = "./data/processed/imputed", pattern="CRC_abundances_\\d*_pc", header = TRUE)
metadata <- read.table("./data/processed/CRC_metadata.txt", header = TRUE)

# merge data_transformed
for (i in 1:length(data_transformed)){
  data_transformed[[i]] <- data_transformed[[i]] %>% 
  left_join(metadata[ ,c("SampleID", "Group", "BMI")], by = "SampleID") %>% 
  relocate(c(Group, BMI), .after = SampleID) %>% 
  select(-c(SampleID, BMI))
}

# merge data_imputed and also split
train_data <- list()
test_data <- list()

for (i in 1:length(data_imputed)){
  
  # merge PCOS_data with metadata
  data_imputed[[i]] <- data_imputed[[i]] %>% 
  left_join(metadata[ ,c("SampleID", "Group", "BMI")], by = "SampleID") %>% 
  relocate(c(Group, BMI), .after = SampleID) %>% 
  select(-c(SampleID, BMI))
  
  # Split the data
  set.seed(2022)
  data_split <- initial_split(data_imputed[[i]], prop = 0.8, strata = Group) 
  train_data[[i]] <- training(data_split) %>% as.data.frame()
  test_data[[i]] <- testing(data_split) %>% as.data.frame()
  
}

names(train_data) <- names(data_imputed)
names(test_data) <- names(data_imputed)

# save integers that control training set
idx <- data_split$in_id
```

mikropml
```{r}
training_AUC <- list()
test_AUC <- list()
counter <- seq(1,10,1)

# go through each data set in data_list
for (i in 1:length(data_transformed)) { 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  # repeat experiment n times
  for (j in counter) { 
    
    ml_result <- run_ml(data_transformed[[i]],
                     'glmnet',
                     outcome_colname = "Group", 
                     kfold = 5,
                     cv_times = 10,
                     training_frac = idx, # training fraction taken from split for codacore
                     seed = NA)
    
    # save training AUC in lists
    training_res[j] <- ml_result$performance[1]
    test_res[j] <- ml_result$performance[3]
  }
  
  # save lists in final list
  training_AUC[[i]] <- training_res
  test_AUC[[i]] <- test_res
}
```

Codacore
```{r}
coda_training_AUC <- list()
coda_test_AUC <- list()

counter <- seq(1,10,1)

for (i in 1:length(train_data)) { 
  
  # generate features (x) and variable (y) for train
  x_train <- train_data[[i]]
  x_train <- x_train[, 2:ncol(x_train)]
  
  y_train <- train_data[[i]]
  y_train <- y_train[,1] 
  
  # generate features (x) and variable (y) for test
  x_test <- test_data[[i]]
  x_test <- x_test[, 2:ncol(x_test)]
  
  y_test <- test_data[[i]]
  y_test <- y_test[,1] 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  for (j in counter) {
    # models
    model_bal_0 <- codacore(
      x_train, 
      y_train, 
      objective = "binary classification",
      logRatioType = "balances",
      cvParams = c(numFolds = 5),
      lambda = 0,
      verbose = FALSE, 
      fast = TRUE
    )
    
    model_bal_1 <- codacore(
      x_train, 
      y_train,
      objective = "binary classification",
      logRatioType = "balances",  
      cvParams = c(numFolds = 5),
      lambda = 1,
      verbose = FALSE, 
      fast = TRUE    
    )
    
    # save training AUC in lists
    training_res[[j]] <- c(as.numeric(model_bal_0$ensemble[[1]]$AUC), 
                         as.numeric(model_bal_1$ensemble[[1]]$AUC))
    
    # get test AUC by predicting manually
    
    
    
    pred_bal_0_lr1 <- predict(model_bal_0, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_0 <- predict(model_bal_0, x_test, logits = FALSE)
    
    pred_bal_1_lr1 <- predict(model_bal_1, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_1 <- predict(model_bal_1, x_test, logits = FALSE)
    
    test_res[[j]] <- c(as.numeric(pROC::auc(y_test, pred_bal_0_lr1, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_0, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_1_lr1, quiet = T)),
                    as.numeric(pROC::auc(y_test, pred_bal_1, quiet = T)))
  }
  
  # save lists in final list
  coda_training_AUC[[i]] <- training_res
  coda_test_AUC[[i]] <- test_res

} 

```

plot mikropml
```{r}
set_names <- names(data_transformed)

names(training_AUC) <- set_names
names(test_AUC) <- set_names

training_df <- stack(training_AUC)
test_df <- stack(test_AUC)

training_df <- training_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "training")

test_df <- test_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "test")
  
final_df_AUC <- rbind(training_df, test_df)

graph <- ggplot(final_df_AUC, aes(x = Dataset, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  labs(title = "AUC over various transformations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graph

ggsave(plot = graph, filename = "./figs/CRC_glmnet_same_split.jpeg", device = "jpeg", height = 6, width = 7)
```

# plot codacore
coda_test_AUC:  1. pred_bal_0_lr1, 2. pred_bal_0, 3. pred_bal_1_lr1; 4. pred_bal_1
coda_training_AUC: 1. model_bal_0 AUC, 2. model_bal_1 AUC

same format as mikropml:
- x-axis: 4 models for each PCOS_10 and PCOS_50 with e.g. model_bal_0 for both pred_bal_0_lr1 and pred_bal_0
```{r}
# set names
set_names <- names(data_imputed)

names(coda_training_AUC) <- set_names
names(coda_test_AUC) <- set_names

# extract first element of coda_training AUC
df_final <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)


# same for PCOS_50
df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC
df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC PCOS_50
df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)


# plot
graph <- ggplot(df_final, aes(x = model, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  facet_wrap(~dataset, scale="free") + 
  labs(title = "AUC CoDaCoRe") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

graph
ggsave(plot = graph, filename = "./figs/CRC_codacore_same_split.jpeg", device = "jpeg", height = 6, width = 7)

```

## CRC predicted on holdout set
```{r}
# Import data
data_transformed <- import(path = "./data/transformed/pseudocount", pattern="CRC_\\d*", header = TRUE)
data_imputed <- import(path = "./data/processed/imputed", pattern="CRC_abundances_\\d*_pc", header = TRUE)
metadata <- read.table("./data/processed/CRC_metadata.txt", header = TRUE)


# merge data_transformed
for (i in 1:length(data_transformed)){
  data_transformed[[i]] <- data_transformed[[i]] %>% 
  left_join(metadata[ ,c("SampleID", "Group", "BMI")], by = "SampleID") %>% 
  relocate(c(Group, BMI), .after = SampleID) %>% 
  dplyr::select(-c(SampleID, BMI))
}


# merge data_imputed and also split
train_data <- list()
test_data <- list()


for (i in 1:length(data_imputed)){
  
  # merge PCOS_data with metadata
  data_imputed[[i]] <- data_imputed[[i]] %>% 
  left_join(metadata[ ,c("SampleID", "Group", "BMI", "Country")], by = "SampleID") %>% 
  relocate(c(Group, BMI, Country), .after = SampleID) %>% 
  dplyr::select(-c(SampleID, BMI))
  
  # Split the data
  train_data[[i]] <- data_imputed[[i]] %>% 
    filter(!grepl("GER", Country)) %>% 
    dplyr::select(-c(Country))
  
  test_data[[i]] <- data_imputed[[i]] %>% 
    filter(grepl("GER", Country)) %>% 
    dplyr::select(-c(Country))  
}

names(train_data) <- names(data_imputed)
names(test_data) <- names(data_imputed)

# save integers that control training set (here holdout)
idx <- data_imputed[[1]] %>%
  rownames_to_column() %>%
  filter(!grepl("GER", Country)) %>%
  `[[`("rowname") %>%
  as.numeric()
```

Pipeline mikropml
```{r}
training_AUC <- list()
test_AUC <- list()
counter <- seq(1,10,1)

# go through each data set in data_list
for (i in 1:length(data_transformed)) { 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  # repeat experiment n times
  for (j in counter) { 
    
    ml_result <- run_ml(data_transformed[[i]],
                     'glmnet',
                     outcome_colname = "Group", 
                     kfold = 5,
                     cv_times = 10,
                     training_frac = idx, # training fraction taken from split for codacore
                     seed = NA)
    
    # save training AUC in lists
    training_res[j] <- ml_result$performance[1]
    test_res[j] <- ml_result$performance[3]
  }
  
  # save lists in final list
  training_AUC[[i]] <- training_res
  test_AUC[[i]] <- test_res
}
```

plot mikropml
```{r}
set_names <- names(data_transformed)

names(training_AUC) <- set_names
names(test_AUC) <- set_names

training_df <- stack(training_AUC)
test_df <- stack(test_AUC)

training_df <- training_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "training")

test_df <- test_df %>% 
  rename(AUC = values, Dataset = ind) %>% 
  add_column(Type = "test")
  
final_df_AUC <- rbind(training_df, test_df)

graph <- ggplot(final_df_AUC, aes(x = Dataset, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  labs(title = "AUC over various transformations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
graph

ggsave(plot = graph, filename = "./figs/CRC_glmnet_same_split_holdout.jpeg", device = "jpeg", height = 6, width = 7)
```
# Pipeline Codacore
```{r}
coda_training_AUC <- list()
coda_test_AUC <- list()

counter <- seq(1,10,1)

for (i in 1:length(train_data)) { 
  
  # generate features (x) and variable (y) for train
  x_train <- train_data[[i]]
  x_train <- x_train[, 2:ncol(x_train)]
  
  y_train <- train_data[[i]]
  y_train <- y_train[,1] 
  
  # generate features (x) and variable (y) for test
  x_test <- test_data[[i]]
  x_test <- x_test[, 2:ncol(x_test)]
  
  y_test <- test_data[[i]]
  y_test <- y_test[,1] 
  
  # initiate temp lists
  training_res <- list()
  test_res <- list()
  
  for (j in counter) {
    # models
    model_bal_0 <- codacore(
      x_train, 
      y_train, 
      objective = "binary classification",
      logRatioType = "balances",
      cvParams = c(numFolds = 5),
      lambda = 0,
      verbose = FALSE, 
      fast = TRUE
    )
    
    model_bal_1 <- codacore(
      x_train, 
      y_train,
      objective = "binary classification",
      logRatioType = "balances",  
      cvParams = c(numFolds = 5),
      lambda = 1,
      verbose = FALSE, 
      fast = TRUE    
    )
    
    # save training AUC in lists
    training_res[[j]] <- c(as.numeric(model_bal_0$ensemble[[1]]$AUC), 
                         as.numeric(model_bal_1$ensemble[[1]]$AUC))
    
    # get test AUC by predicting manually
    
    
    
    pred_bal_0_lr1 <- predict(model_bal_0, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_0 <- predict(model_bal_0, x_test, logits = FALSE)
    
    pred_bal_1_lr1 <- predict(model_bal_1, x_test, logits = FALSE, numLogRatios = 1)
    pred_bal_1 <- predict(model_bal_1, x_test, logits = FALSE)
    
    test_res[[j]] <- c(as.numeric(pROC::auc(y_test, pred_bal_0_lr1, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_0, quiet = T)), 
                    as.numeric(pROC::auc(y_test, pred_bal_1_lr1, quiet = T)),
                    as.numeric(pROC::auc(y_test, pred_bal_1, quiet = T)))
  }
  
  # save lists in final list
  coda_training_AUC[[i]] <- training_res
  coda_test_AUC[[i]] <- test_res

} 

```

# plot codacore
```{r}
# set names
set_names <- names(data_imputed)

names(coda_training_AUC) <- set_names
names(coda_test_AUC) <- set_names

# extract first element of coda_training AUC
df_final <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df <- data.frame(AUC = sapply(coda_training_AUC[[1]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)


# same for PCOS_50
df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",1),
                 model = "model_bal_0",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_training_AUC[[2]],"[[",2),
                 model = "model_bal_1",
                 Type = "training",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC
df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[1]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_10_pc")

df_final <- rbind(df_final, df)

# same for coda_test_AUC PCOS_50
df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",1),
                 model = "pred_bal_0_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",2),
                 model = "pred_bal_0",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",3),
                 model = "pred_bal_1_lr1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)

df <- data.frame(AUC = sapply(coda_test_AUC[[2]],"[[",4),
                 model = "pred_bal_1",
                 Type = "test",
                 dataset = "CRC_abundances_50_pc")

df_final <- rbind(df_final, df)


# plot
graph <- ggplot(df_final, aes(x = model, y = AUC, fill = Type)) + 
  geom_boxplot() + 
  facet_wrap(~dataset, scale="free") + 
  labs(title = "AUC CoDaCoRe") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

graph
ggsave(plot = graph, filename = "./figs/CRC_codacore_holdout.jpeg", device = "jpeg", height = 6, width = 7)
```