---
title: "1_pre_processing_data"
author: "Jennifer Neumaier"
date: "2/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pre-Processing of Data

```{r}
setwd("D:/Master/Master Biologie/Sommersemester 2022/Master thesis/Github")
```


Importing libraries
```{r}
# Filtering
library("tidyverse")

# Zero-Shot Imputation
library("zCompositions")
```

Importing source files
```{r}
# source("./src/0_get_raw_data.R")
source("./src/data_analysis.R")
source("./src/convenience.R")
```


## EstMB

Processing EstMB metadata
```{r}
EstMB_abundances <- EstMB_abundances %>% 
  rename("SampleID" = skood)

EstMB_metadata <- EstMB_metadata %>% 
  rename("SampleID" = skood)

EstMB_metadata <- EstMB_metadata %>% 
  drop_na(F41, I11, E11, BMI, Age_at_MBsample)

# cut patients from EstMB_abundances that are not in metadata
diff <- setdiff(EstMB_abundances$SampleID, EstMB_metadata$SampleID)
EstMB_abundances <- EstMB_abundances[!EstMB_abundances$SampleID %in% diff,]

write.table(EstMB_metadata,"./data/processed/EstMB_metadata.txt", row.names = FALSE)

write.table(EstMB_abundances,"./data/processed/EstMB_abundances.txt", row.names = FALSE)
```

Filtering data set
Two pre-processed versions will be created:
1. Filtering out all species that account for < 10% abundance in patients.
2. Filtering out all species that account for < 50% abundance in patients.
3. Add mean relative abundance filter with at least 0.1% mean relative abundance.
```{r}
# percent.filter: 10% -> 0.10
# relabund: 0.1% mean -> 0.001
EstMB_abundances_10 <- taxa.filter(EstMB_abundances, percent.filter = 0.10, relabund.filter = 0.001)
write.table(EstMB_abundances_10,"./data/processed/filtered/EstMB_abundances_10.txt", row.names = FALSE)

EstMB_abundances_50 <- taxa.filter(EstMB_abundances, percent.filter = 0.50, relabund.filter = 0.001)
write.table(EstMB_abundances_50,"./data/processed/filtered/EstMB_abundances_50.txt", row.names = FALSE)
```

Zero-Imputation EstMB
```{r}
# pseudocount (pc)
EstMB_abundances_10_pc <- cbind(EstMB_abundances_10[1], cmultRepl(EstMB_abundances_10[,2:ncol(EstMB_abundances_10)], output = "p-counts"))
EstMB_abundances_10_pc <- EstMB_abundances_10_pc %>% 
  mutate_if(is.numeric, round, digits=3)

write.table(EstMB_abundances_10_pc,"./data/processed/imputed/EstMB_abundances_10_pc.txt", row.names = FALSE)


EstMB_abundances_50_pc <- cbind(EstMB_abundances_50[1], cmultRepl(EstMB_abundances_50[,2:ncol(EstMB_abundances_50)], output = "p-counts"))
EstMB_abundances_50_pc <- EstMB_abundances_50_pc %>% 
  mutate_if(is.numeric, round, digits=3)

write.table(EstMB_abundances_50_pc,"./data/processed/imputed/EstMB_abundances_50_pc.txt", row.names = FALSE)
```

mbImpute
Should ideally be used with phylogenetic information. Metadata is optional. As in this case phylogenetic information is not available, we have to impute only one condition.
```{r}
# mbImpute (mb)
```


## PCOS

Processing PCOS metadata and cutting out non-sequenced samples
```{r}
# Transposing data frame and renaming
PCOS_abundances <- data.frame(t(PCOS_abundances))

# fit PCOS tables to EstMB design
colnames(PCOS_abundances) <- PCOS_abundances[1,]
PCOS_abundances <- PCOS_abundances[-1,]

PCOS_abundances <- PCOS_abundances %>%
  rownames_to_column(var="SampleID")

# Keeping only phenotype column of metadata (PCOS_Riikka) and BMI (C6646C_weigth_height_007)
PCOS_metadata <- PCOS_metadata[,c("SampleID", "PCOS_Riikka", "C6646C_weigth_height_007")]
PCOS_metadata <- PCOS_metadata %>% 
  rename(BMI = C6646C_weigth_height_007) 
write.table(PCOS_metadata,"./data/processed/PCOS_metadata.txt", row.names = FALSE)

# cut patients from PCOS_abundances that are not in metadata
diff <- setdiff(PCOS_abundances$SampleID, PCOS_metadata$SampleID)
PCOS_abundances <- PCOS_abundances[!PCOS_abundances$SampleID %in% diff,]

# convert all entries to numeric
PCOS_abundances <- PCOS_abundances %>% 
  mutate_at(vars(-("SampleID")),function(x) as.numeric(x))

write.table(PCOS_abundances,"./data/processed/PCOS_abundances.txt", row.names = FALSE)
```

Filtering data set
Two pre-processed versions will be created:
1. Filtering out all species that account for < 10% abundance in patients.
2. Filtering out all species that account for < 50% abundance in patients.
3. Add mean relative abundance filter with at least 0.1% mean relative abundance.
```{r}
# percent.filter: 10% -> 0.10
# relabund: 0.1% mean -> 0.1
PCOS_abundances_10 <- taxa.filter(PCOS_abundances, percent.filter = 0.10, relabund.filter = 0.001)
write.table(PCOS_abundances_10,"./data/processed/filtered/PCOS_abundances_10.txt", row.names = FALSE)

PCOS_abundances_50 <- taxa.filter(PCOS_abundances, percent.filter = 0.50, relabund.filter = 0.001)
write.table(PCOS_abundances_50,"./data/processed/filtered/PCOS_abundances_50.txt", row.names = FALSE)
```

Zero-Imputation PCOS
```{r}
# pseudocount (pc)
# use zCompositions, round all values and if they are smaller than 0, replace with 1
PCOS_abundances_10_pc <- cbind(PCOS_abundances_10[1], cmultRepl(PCOS_abundances_10[,2:ncol(PCOS_abundances_10)], output = "p-counts"))
PCOS_abundances_10_pc <- PCOS_abundances_10_pc %>% 
  mutate_if(is.numeric, round, digits=3)

write.table(PCOS_abundances_10_pc,"./data/processed/imputed/PCOS_abundances_10_pc.txt", row.names = FALSE)


PCOS_abundances_50_pc <- cbind(PCOS_abundances_50[1], cmultRepl(PCOS_abundances_50[,2:ncol(PCOS_abundances_50)], output = "p-counts"))
PCOS_abundances_50_pc <- PCOS_abundances_50_pc %>% 
  mutate_if(is.numeric, round, digits=3)

write.table(PCOS_abundances_50_pc,"./data/processed/imputed/PCOS_abundances_50_pc.txt", row.names = FALSE)
```

mbImpute
Should ideally be used with phylogenetic information. Metadata is optional. As in this case phylogenetic information is not available, we have to impute only one condition.
```{r}
# mbImpute (mb)
```


## CRC
Processing CRC data set and cutting out non-sequenced samples

needs work -> too many samples in abundance table compared to metadata -> cut out here
```{r}
# rename Sample_ID column and replace all - with . in SampleIDs. 
names(CRC_metadata)[names(CRC_metadata) == "Sample_ID"] <- "SampleID"
CRC_metadata$SampleID <- chartr("-", ".", CRC_metadata$SampleID)

# delete rows where BMI is missing and keep one data set back as holdout
CRC_metadata <- CRC_metadata %>% 
  drop_na(BMI, Group)

write.table(CRC_metadata,"./data/processed/CRC_metadata.txt", row.names = FALSE)

# Transposing data frame and renaming
CRC_abundances <- data.frame(t(CRC_abundances))

# make extra column SampleID
CRC_abundances <- cbind(SampleID = rownames(CRC_abundances), CRC_abundances)
rownames(CRC_abundances) <- 1:nrow(CRC_abundances)

# delete rows in abundance table that were removed from metadata
diff <- setdiff(CRC_abundances$SampleID, CRC_metadata$SampleID)
CRC_abundances <- CRC_abundances[!CRC_abundances$SampleID %in% diff,]

write.table(CRC_abundances,"./data/processed/CRC_abundances.txt", row.names = FALSE)
```

Filtering data set
Two pre-processed versions will be created:
1. Filtering out all species that account for < 10% abundance in patients.
2. Filtering out all species that account for < 50% abundance in patients.
3. Add mean relative abundance filter with at least 0.1% mean relative abundance.
```{r}
# Filtering
# percent.filter: 10% -> 0.10
# relabund: 0.1% mean -> 0.001
CRC_abundances_10 <- taxa.filter(CRC_abundances, percent.filter = 0.10, relabund.filter = 0.001)
write.table(CRC_abundances_10,"./data/processed/filtered/CRC_abundances_10.txt", row.names = FALSE)

CRC_abundances_50 <- taxa.filter(CRC_abundances, percent.filter = 0.50, relabund.filter = 0.001)
write.table(CRC_abundances_50,"./data/processed/filtered/CRC_abundances_50.txt", row.names = FALSE)
```

Zero-Imputation CRC
```{r}
# pseudocount (pc)
CRC_abundances_10_pc <- cbind(CRC_abundances_10[1], cmultRepl(CRC_abundances_10[,2:ncol(CRC_abundances_10)], output = "p-counts"))
CRC_abundances_10_pc <- CRC_abundances_10_pc %>% 
  mutate_if(is.numeric, round, digits=3)

write.table(CRC_abundances_10_pc,"./data/processed/imputed/CRC_abundances_10_pc.txt", row.names = FALSE)


CRC_abundances_50_pc <- cbind(CRC_abundances_50[1], cmultRepl(CRC_abundances_50[,2:ncol(CRC_abundances_50)], output = "p-counts"))
CRC_abundances_50_pc <- CRC_abundances_50_pc %>% 
  mutate_if(is.numeric, round, digits=3)


write.table(CRC_abundances_50_pc,"./data/processed/imputed/CRC_abundances_50_pc.txt", row.names = FALSE)
```

